---
title: "Data Cleaning & Manipulation"
author: "Halina Do-Linh"
date: "2023-03-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries

```{r}
library(dplyr)
library(tidyr)
library(readr)
```

## Load data

```{r}
catch_original <- read.csv("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.302.1")
```

## Clean data

`select()` from `dplyr` package selects **columns**.

```{r select cols}
# remove all notesRegCode columns from the catch_original data frame
catch_data <- catch_original %>% 
  select(-All, -notesRegCode)
```

By exploring the data we see that some variables have classes we do not expect.

```{r explore data}
# shows first 6 rows of df
head(catch_data)

# same as head but different view
glimpse(catch_data)
```
This warning tells us that somewhere in the dataset, the `Chinook` col has a character or a `NA` value when it should have a number

Show in New Window
Warning: There was 1 warning in `mutate()`.
ℹ In argument: `Chinook = as.numeric(Chinook)`.
Caused by warning:
! NAs introduced by coercion

```{r change class}
catch_clean <- catch_data %>% 
  mutate(Chinook = as.numeric(Chinook))
```
This tells me that my `NA` value is located at row 401.

```{r find NA}
i <- which(is.na(catch_clean$Chinook))

# I is not a 1!
catch_data[i,]
```

```{r convert "I" to 1}
# note we are assuming that the "I" is a typo vs a coded missing value, etc
# always check the metadata
catch_clean <- catch_data %>% 
  mutate(Chinook = if_else(Chinook == "I", "1", Chinook)) %>% 
  mutate(Chinook = as.integer(Chinook))

glimpse(catch_clean)
```

## Halina Notes
- `pivot_longer()` the `names_to` arg is taking all the names of the columns and turning it into a new column where the names turn into the values for that new col you're creating. The `values_to` arg is taking all the values in from those original columns are placed into a new col that we've named in `values_to`.

```{r pivot longer}
catch_long <- catch_clean %>% 
  pivot_longer(cols = -c(Region, Year),
               names_to = "species",
               values_to = "catch")
```

```{r pivot wider}
catch_wide <- catch_long %>% 
  pivot_wider(names_from = species,
              values_from = catch)
```

## Practice renaming columns

```{r}
catch_long <- catch_long %>% 
  rename(catch_thousands = catch)
```

## Add columns with `mutate()`

```{r change catch metric}
catch_long <- catch_long %>% 
  mutate(catch = catch_thousands * 1000)

# remove `catch_thousands`
catch_long <- catch_long %>% 
  select(-catch_thousands)
```
 
## Pratice using `group_by()` and `summarize()`

```{r mean counts by region}
mean_region <- catch_long %>% 
  group_by(Region) %>% 
  summarize(catch_mean = mean(catch))
```

```{r number of obs by region}
n_region <- catch_long %>% 
  group_by(Region) %>% 
  summarize(n = n(), # number of rows of each group
            total_catch = sum(catch)) 
```

## Filtering

`filter()` from `dplyr` package filters **rows**.

```{r}
ALU_catch <- catch_long %>%
  # multiple regions
  # use %in% over == because == will be looking for the pattern ALU (first) SSE (second)
  # filter(Region %in% c("ALU", "SSE")) %>% 
  filter(Region == "ALU")

```

## Arrange some rows

```{r}

```

























