---
title: "Geospatial Analysis with sf"
author: "Halina Do-Linh"
date: "2023-03-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setup

```{r load libraries}
library(sf)
library(ggplot2)
library(ggmap)
library(scales)
library(leaflet)
library(dplyr)
```

```{r read data}
ak_regions <- read_sf("data/ak_regions_simp.shp")
```

```{r}
plot(ak_regions)
```

```{r class ak_regions}
class(ak_regions)
```

```{r}
head(ak_regions)
```



## Coordinate Reference Systems (CRS)

A CRS can be both unprojected and projected.

Unprojected CRS aka a geographic coordinate system (GCS)

- latitude and longitude references the earth as a 3D object
- tells us *where* something is located on the globe
- measured in angular units, i.e. degrees

Projected CRS or projected coordinated system (PCS) or projection (for short)

- latitude and longitude references the earth as a 2D object
- tells us *how* to draw or create the representation on a flat surface
- measured in linear units, i.e. meters

```{r find crs}
st_crs(ak_regions)
```

```{r transform crs}
ak_regions_3338 <- ak_regions %>% 
  st_transform(crs = 3338)

st_crs(ak_regions_3338)
```

```{r plot with ak albers crs}
plot(ak_regions_3338)
```

## sf & the tidyverse

```{r}
ak_regions_3338 %>% 
  select(region)
```

```{r}
ak_regions_3338 %>% 
  filter(region == "Southeast")
```

## spatial joins

```{r read in pop data}
pop <- read.csv("data/alaska_population.csv")

class(pop)
```

### make `pop` an sf object using `st_as_sf()`

```{r}
pop_4326 <- st_as_sf(pop,
                     coords = c("lng", "lat"),
                     crs = 4326,
                     remove = FALSE)

head(pop_4326)
```
## assign each city to a region using `st_join()`

```{r}
pop_joined <- st_join(pop_4326, ak_regions_3338, join = st_within)
```

```{r}
pop_3338 <- st_transform(pop_4326, crs = 3338)

st_crs(pop_3338)
```

```{r}
pop_joined <- st_join(pop_3338, ak_regions_3338, join = st_within)

head(pop_joined)
```

## calculate the total population per region using `group_by()` and `summarize()`

```{r}
pop_region <- pop_joined %>% 
  as.data.frame() %>% 
  group_by(region) %>% 
  summarize(total_pop = sum(population))

head(pop_region)
```

```{r}
pop_region_3338 <- left_join(ak_regions_3338, pop_region, by = "region")

head(pop_region_3338)
```

```{r}
plot(pop_region_3338["total_pop"])
```

```{r}
pop_mgmt_3338 <- pop_region_3338 %>% 
  group_by(mgmt_area) %>% 
  summarize(total_pop = sum(total_pop), do_union = FALSE)

plot(pop_mgmt_3338["total_pop"])
```

## Writing spatial data as a file using `write_sf()`

```{r}
write_sf(pop_region_3338, "data/ak_regions_populations.shp", delete_layer = TRUE)
```

## Visualize using `ggplot()`

```{r}
ggplot(pop_region_3338) +
  geom_sf(aes(fill = total_pop)) +
  theme_bw() +
  labs(fill = "Total Popualation") +
  scale_fill_continuous(low = "khaki", high = "firebrick", labels = comma) # comma comes from scales
```

```{r read rivers shp}
rivers_3338 <- read_sf("data/ak_rivers_simp.shp")

st_crs(rivers_3338)
```

```{r}
ggplot() +
  geom_sf(data = pop_region_3338, aes(fill = total_pop)) +
  geom_sf(data = rivers_3338, aes(size = StrOrder), color = "black") +
  geom_sf(data = pop_3338, size = 0.5) +
  scale_size(range = c(0.01, 0.2), guide = F) +
  theme_bw() +
  labs(fill = "Total Population") +
  scale_fill_continuous(low = "khaki", high = "firebrick", labels = comma)
```

## Incorporate base maps into static maps using `ggmamp`

```{r}
pop_3857 <- pop_3338 %>% 
  st_transform(crs = 3857)
```

```{r}
# Define a function to fix the bbox to be in EPSG:3857
# See https://github.com/dkahle/ggmap/issues/160#issuecomment-397055208
ggmap_bbox_to_3857 <- function(map) {
  if (!inherits(map, "ggmap")) stop("map must be a ggmap object")
  # Extract the bounding box (in lat/lon) from the ggmap to a numeric vector, 
  # and set the names to what sf::st_bbox expects:
  map_bbox <- setNames(unlist(attr(map, "bb")), 
                       c("ymin", "xmin", "ymax", "xmax"))
  
  # Coonvert the bbox to an sf polygon, transform it to 3857, 
  # and convert back to a bbox (convoluted, but it works)
  bbox_3857 <- st_bbox(st_transform(st_as_sfc(st_bbox(map_bbox, crs = 4326)), 3857))
  
  # Overwrite the bbox of the ggmap object with the transformed coordinates 
  attr(map, "bb")$ll.lat <- bbox_3857["ymin"]
  attr(map, "bb")$ll.lon <- bbox_3857["xmin"]
  attr(map, "bb")$ur.lat <- bbox_3857["ymax"]
  attr(map, "bb")$ur.lon <- bbox_3857["xmax"]
  map
}
```

```{r}
bbox <- c(-170, 52, -130, 64) # roughly southern Alaska
ak_map <- get_stamenmap(bbox, zoom = 4)
ak_map_3857 <- ggmap_bbox_to_3857(ak_map)
```

```{r}
ggmap(ak_map_3857) +
  geom_sf(data = pop_3857, aes(color = population), inherit.aes = F) +
  scale_color_continuous(low = "khaki", high = "firebrick", labels = comma)
```

## Interactive maps with `leaflet`

```{r}
epsg3338 <- leaflet::leafletCRS(
  crsClass = "L.Proj.CRS",
  code = "EPSG:3338",
  proj4def =  "+proj=aea +lat_1=55 +lat_2=65 +lat_0=50 +lon_0=-154 +x_0=0 +y_0=0 +ellps=GRS80 +towgs84=0,0,0,0,0,0,0 +units=m +no_defs",
  resolutions = 2^(16:7))
```

```{r}
pop_region_4326 <- pop_region_3338 %>% 
  st_transform(crs = 4326)
```

```{r}
m <- leaflet(options = leafletOptions(crs = epsg3338)) %>%
        addPolygons(data = pop_region_4326, 
                    fillColor = "gray",
                    weight = 1)

m
```

```{r}
pal <- colorNumeric(palette = "Reds", domain = pop_region_4326$total_pop)

m <- leaflet(options = leafletOptions(crs = epsg3338)) %>%
        addPolygons(data = pop_region_4326, 
                    fillColor = ~pal(total_pop),
                    weight = 1,
                    color = "black",
                    fillOpacity = 1) %>% 
        addCircleMarkers(data = pop_4326,
                         lat = ~lat,
                         lng = ~lng,
                         radius = ~log(population/500), # arbitrary scaling
                         fillColor = "gray",
                         fillOpacity = 1,
                         weight = 0.25,
                         color = "black",
                         label = ~paste0(pop_4326$city, ", population ", comma(pop_4326$population))) %>%
        addLegend(position = "bottomleft",
                  pal = pal,
                  values = range(pop_region_4326$total_pop),
                  title = "Total Population")

m
```

































