---
title: "Practice RMarkdown"
author: "Halina Do-Linh"
date: "2023-02-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(readr)
```

## Introduction

This dataset has many observations of measurements of seawater chemistry. It is called the North Pole Environmental Observatory Bottle Chemistry data. [Learn more about the data and download it online from its Arctic Data Center data repository page](https://arcticdata.io/catalog/view/doi:10.18739/A25T3FZ8X).

### Read in data

```{r, message=FALSE}
bg_chem <- read_csv("data/BGchem2008data.csv")
```

```{r exploration, echo=FALSE, eval=FALSE}
colnames(bg_chem)

str(bg_chem)

head(bg_chem)
```

## Analysis

### Calculate summary statistics

```{r summary stats}
nitrate <- mean(bg_chem$NO3)
nitrite <- mean(bg_chem$NO2)
amm <- mean(bg_chem$NH4)
phos <- mean(bg_chem$P)
```

### Calculate mean Redfield ratio

```{r ratio}
ratio <- (nitrate + nitrite + amm)/phos
```

The Redfield ratio is approximately `r round(ratio)`.

### Plot

```{r plot, echo=FALSE}
plot(bg_chem$P, bg_chem$NO2 + bg_chem$NO3 + bg_chem$NH4)
```

